---
title: "edgeR"
author: "Katarzyna Sikora"
output: html_document
---

```{r setup, include=FALSE}
wd<-file.path(snakemake@params[["basedir"]],"edgeR_output")
system(paste0('mkdir -p ',wd))
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = wd)
```

## Libs

```{r libs}
library(tximport)
library(edgeR)
library(GenomicRanges)
```

## Catch Salmon

Oarfish output was reformatted to match Salmon output.

The Salmon-like output directories are read using edgeR’s catchSalmon function, which imports the transcript-level counts and estimates the overdispersion associated with read-to-transcript assignment ambiguity.

```{r catch salmon}
input_files<-unlist(snakemake@params[["input_files"]])
dirs<-dirname(input_files)

data<-catchSalmon(dirs)

```

## Prep the DGEList object

To account for the mapping ambiguity resulting from the assignment of reads to transcripts, we use the divided count strategy that divides the transcript-level counts by the estimated mapping ambiguity overdispersion associated to each transcript.

For reference, see DOI: 10.1093/nar/gkad1167 and 10.1093/nargab/lqae151 .

```{r sampleinfo}
colnames(data$counts)<-basename(colnames(data$counts))
divided.counts <- data$counts/data$annotation$Overdispersion

#write.table(divided.counts,"divided.counts.tsv",sep="\t")

sampleSheet<-snakemake@input[["sampleSheet"]]
sampleInfo<-read.table(sampleSheet,sep="\t")
rownames(sampleInfo)<-sampleInfo$name
divided.counts<-divided.counts[,match(sampleInfo$name,colnames(divided.counts)]


y <- DGEList(counts = divided.counts, samples = sampleInfo, group = sampleInfo$condition, genes = data$annotation)

head(y$genes)

gtf_file<-snakemake@input[["gtf_file"]]


gtf<-rtracklayer::import(gtf_file)

ginfo <- mcols(gtf)[match(rownames(y),mcols(gtf)$transcript_id),c("transcript_type","gene_id","gene_name")]
y$genes <- cbind(y$genes,ginfo)
head(y$genes)
```


## Filter and normalize

Lowly expressed transcripts are filtered out prior to the downstream analysis.   

After dividing counts by the estimated mapping ambiguity overdispersion, scaling factors can computed using the TMM method to convert the resulting library sizes to effective library sizes.   

```{r filt norm}
keep <- filterByExpr(y)
table(keep)
keep
y <- y[keep, , keep.lib.sizes=FALSE]


y <- normLibSizes(y)
y$samples
```


## Calculate MDS

MDS plots can also be used to visualize differences between the expression profiles of different samples with transcript-level counts.   

```{r MDS}
plotMDS(y,col = c(1:2)[y$samples$condition],labels = y$samples$name,xlim = c(-4,4))
```

## Design matrix

We create the design matrix to compare HEK293 cells against HAP1 cells.   

```{r design}
design <- model.matrix(~ condition, data = y$samples)
design
```


## Dispersion estimation

Estimate and visualize NB dispersions.   

```{r disp}
y <- estimateDisp(y, design, robust=TRUE)
y$common.dispersion
plotBCV(y)

```

The NB dispersion estimates will not be used further under the latest quasi-likelihood (QL) pipeline.   
For DTE analyses, we still recommend using the quasi-likelihood (QL) pipeline for stricter error rate control by accounting for the uncertainty associated with the dispersion estimation.

```{r quasil}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```

## Differential expression

Differentially expressed transcripts are tested between cell lines using the QL F-test.

```{r dte}
qlf <- glmQLFTest(fit)
is.de <- decideTests(qlf, p.value=0.05)
summary(is.de)


tt <- as.data.frame(topTags(qlf, n = Inf,p.value=0.05))
head(tt)
table(tt$transcript_type)
length(unique(tt$gene_id))
```
## MA plot

```{r ma}
plotMD(qlf)
```

## Save and export results

```{r save}
saveRDS(qlf,"qlf.RDS")
write.table(tt,"topTags_pval0.05.tsv",sep="\t",quote=FALSE)

```


## Session Info

```{r session info}
sessionInfo()
```


